Fusing Filters with Integer Linear Programming

### 2014-05-16 2014betterfusionforfilters/section/03-SizeInference.tex
note on transitivity. make consistent section and subsection capitalisation

%!TEX root = ../Main.tex
\section{Size inference}
Before performing fusion proper, we must infer the relative sizes of each array in the program. We achieve this with a simple constraint based inference algorithm, which we discuss in this section. Size inference has been previously described in the context of array fusion by Chatterjee~\cite{chatterjee1991size}. In constrast to our algorithm, \cite{chatterjee1991size} does not support size changing functions such as filter.

@@

% -----------------------------------------------------------------------------
\subsection{Size types, constraints and schemes}
\label{s:SizeTypes}
Figure~\ref{f:constraints} shows the grammar for size types, constraints and schemes. A size scheme is like a type constraint from Hindley-Milner type systems, except that it only mentions the size of each input array, instead of the element types as well.

@@

% -----------------------------------------------------------------------------
\subsection{Constraint Generation}
The rules for constraint generation are shown in Figure~\ref{f:ConstraintGeneration}. The top level judgment ~~$\SizeF{function}{\sigma}$~~ assigns a size scheme to a function. It does this by extracting size constraints and then solving them. This rule, along with the constraint solving process is discussed in the next section. The judgment ~~$\SizeB{\Gamma_1}{zs}{b}{\Gamma_2}{C}$ reads: ``Under environment~$\Gamma_1$, array variable $zs$ binds the result of $b$, producing a result environment $\Gamma_2$ and size constraints $C$''. The remaining judgment that extracts constraints from a list of bindings is similar. The environment $\Gamma$ has the following grammar:
$$
\Gamma~ @::=@ ~~\cdot ~~|~~ \Gamma,~ \Gamma ~~|~~ zs : k ~~|~~ k ~~|~~ \exists k
$$

@@

% -----------------------------------------------------------------------------
\subsection{Iteration size}
After inferring the size of each array variable, each operator is assigned an \emph{iteration size}, which is the number of iterations needed in the loop which evaluates that operator. For @filter@ and other size changing operators, the iteration and result sizes are in general different. For such an operator, we say that the result size is a \emph{descendant} of the iteration size. Conversely, the iteration size is a \emph{parent} of the result size.

>>>>

%!TEX root = ../Main.tex
\section{Size Inference}
Before performing fusion proper, we must infer the relative sizes of each array in the program. We achieve this with a simple constraint based inference algorithm, which we discuss in this section. Size inference has been previously described in the context of array fusion by Chatterjee~\cite{chatterjee1991size}. In constrast to our algorithm, \cite{chatterjee1991size} does not support size changing functions such as filter.

@@

% -----------------------------------------------------------------------------
\subsection{Size Types, Constraints and Schemes}
\label{s:SizeTypes}
Figure~\ref{f:constraints} shows the grammar for size types, constraints and schemes. A size scheme is like a type constraint from Hindley-Milner type systems, except that it only mentions the size of each input array, instead of the element types as well.

@@

% -----------------------------------------------------------------------------
\subsection{Constraint Generation}
The rules for constraint generation are shown in Figure~\ref{f:ConstraintGeneration}. The top level judgment ~~$\SizeF{\function}{\sigma}$~~ assigns a size scheme to a function. It does this by extracting size constraints and then solving them. This rule, along with the constraint solving process is discussed in the next section. The judgment ~~$\SizeB{\Gamma_1}{zs}{b}{\Gamma_2}{C}$ reads: ``Under environment~$\Gamma_1$, array variable $zs$ binds the result of $b$, producing a result environment $\Gamma_2$ and size constraints $C$''. The remaining judgment that extracts constraints from a list of bindings is similar. The environment $\Gamma$ has the following grammar:
$$
\Gamma~ @::=@ ~~\cdot ~~|~~ \Gamma,~ \Gamma ~~|~~ zs : k ~~|~~ k ~~|~~ \exists k
$$

@@

% -----------------------------------------------------------------------------
\subsection{Iteration Size}
After inferring the size of each array variable, each operator is assigned an \emph{iteration size}, which is the number of iterations needed in the loop which evaluates that operator. For @filter@ and other size changing operators, the iteration and result sizes are in general different. For such an operator, we say that the result size is a \emph{descendant} of the iteration size. Conversely, the iteration size is a \emph{parent} of the result size.

### 2014-05-16 2014betterfusionforfilters/section/04-ILP.tex
note on transitivity. make consistent section and subsection capitalisation

% -----------------------------------------------------------------------------
\subsection{The objective function}
\label{s:ObjectiveFunction}

@@

% -----------------------------------------------------------------------------
\subsection{A note on transitivity}
It may seem that we could generate far fewer constraints and rely on transitivity of clustering equality $x_{ij}$.
This means that for each pair $i,j$ we wouldn't generate an $x_{ij}$ and thus not generate those constraints, unless $i$ or $j$ is a @filter@, or arc between $i$ and $j$.
This would help, but also means we can't \emph{minimise} on these removed $x_{ij}$, so the solution won't count the (rather smaller) benefits of fusing two non-connected nodes.

% -----------------------------------------------------------------------------
\subsection{Fusion-preventing path optimisation}
\label{s:OptimisedConstraints}
We can remove certain constraints and variables, by only checking nodes if there is no \emph{fusion-preventing path} between $i$ and $j$.
If there is a fusion-preventing path between two nodes, they are known to be in different clusters.
With our specific knowledge of the problem domain, we can simplify the integer linear program and make the job of the ILP solver easier by omitting these constraints.
Creating fewer constraints and variables should make it faster for the ILP solver to find a solution.

>>>>

% -----------------------------------------------------------------------------
\subsection{The Objective Function}
\label{s:ObjectiveFunction}

@@

% -----------------------------------------------------------------------------
\subsection{Transitivity of Clustering}
It may seem that we could generate far fewer constraints and rely on transitivity of clustering equality $x_{ij}$.
This means that for each pair $i,j$ we would only generate an $x_{ij}$ and constraints, if $i$ or $j$ is a @filter@, or there is an edge between $i$ and $j$.
This would simplify the integer linear program, but would also modify the objective function not to \emph{minimise} on these removed $x_{ij}$.
The objective function of this simplified version would not count the benefits of fusing two non-connected nodes, and would not reduce memory traffic as effectively.

% -----------------------------------------------------------------------------
\subsection{Fusion-preventing Path Optimisation}
\label{s:OptimisedConstraints}
We can remove certain constraints and variables, by only checking nodes if there is no \emph{fusion-preventing path} between $i$ and $j$.
If there is a fusion-preventing path between two nodes, they are known to be in different clusters.
With our specific knowledge of the problem domain, we can simplify the integer linear program and make the job of the ILP solver easier by omitting these constraints.
Creating fewer constraints and variables should make it faster for the ILP solver to find a solution.

### 2014-05-16 2014betterfusionforfilters/section/06-Related.tex
note on transitivity. make consistent section and subsection capitalisation

%!TEX root = ../Main.tex
\section{Related work}
The idea of using integer linear programming to cluster an operator graph for array fusion was first fully described by Megiddo and Sarkar~\cite{megiddo1998optimal}~(1999). This algorithm was then extended to support multiple iteration sizes by Darte and Huard~\cite{darte2002contraction}. Both algorithms were developed in the context of imperative languages (Fortran) and are based around a Loop Dependence Graph (LDG). In a LDG the nodes represent imperative loops, and the edges indicate which loops may or may not be fused. Although this work was developed in a context of imperative programming, the conceptual framework and algorithms are language agnostic. In earlier work, Chatterjee~\cite{chatterjee1993nested} (1991) mentioned that ILP can be used to schedule a data flow graph, though did not give a complete formulation. Our system extends the prior ILP approaches with support for size changing operators such as @filter@.

>>>>

%!TEX root = ../Main.tex
\section{Related Work}
The idea of using integer linear programming to cluster an operator graph for array fusion was first fully described by Megiddo and Sarkar~\cite{megiddo1998optimal}~(1999). This algorithm was then extended to support multiple iteration sizes by Darte and Huard~\cite{darte2002contraction}. Both algorithms were developed in the context of imperative languages (Fortran) and are based around a Loop Dependence Graph (LDG). In a LDG the nodes represent imperative loops, and the edges indicate which loops may or may not be fused. Although this work was developed in a context of imperative programming, the conceptual framework and algorithms are language agnostic. In earlier work, Chatterjee~\cite{chatterjee1993nested} (1991) mentioned that ILP can be used to schedule a data flow graph, though did not give a complete formulation. Our system extends the prior ILP approaches with support for size changing operators such as @filter@.
